---
title: "Flatbed Scans"
author: "Johannes Cunow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flatbed Scans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This vignette demonstrates how to analyze flatbed scans using the **RootScape Package**. The workflow is streamlined as flatbed scans do not require depth information or calibration, focusing on skeletonization and basic root trait extraction.

## Installation

To install the package from the source or GitHub:

```{r}
# Install the package from GitHub
# install.packages("remotes")
# remotes::install_github("jcunow/RootScape")

# install.packages("devtools")
# devtools::install_github("jcunow/RootScape")

# Load the package
library(RootScape)
```

## Key Functionalities

This route includes:

1. Skeletonization of flatbed scans.
2. Root trait extraction (e.g., root length, root pixel sum).
3. Extra features like peat texture and rhizosphere analysis.

## Example Workflow

### 1. Load Image & setup:

```{r}
library(RootScape)
library(tidyverse)

# load image
img = load_flexible_image("path/image.tiff")
```


### 2. Skeletonization

Skeletonize the flatbed scan:

```{r}
skeleton <- skeletonize_image(img, 
                              binarize = TRUE, 
                              output_format = "spatrast")
plot(skeleton, main = "Skeletonized Image")
```

### 3. Root Trait Extraction

Extract key root traits:
The RootScape package offers a couple of root trait estimations. 

```{r}
# Extract Brancnhing Points
SkeletonPoints <- detect_skeleton_points(skeleton)
roottips = px.sum(SkeletonPoints$Endpoints)
branchingpoints = px.sum(SkeletonPoints$Branchingoints)

# Get Kimura Root Length
rootlength = RootLength(skeleton)

# Calculate Root Cover Density
void = (load_flexible_image(img, binarize = TRUE)-1)*-1
rootpx = px.sum(img, select.layer = 2)
voidpx = px.sum(void, select.layer = 2)
rootdensity = rootpx / (rootpx + voidpx)

# Get the 'deepdrive' trait
ddrive = deep.drive(RootMap = img, DepthMap = DepthMap)

# diameter estimation
diameters = root.diameters(img, skeleton_method = "MAT", select.layer = 2)

# a couple of others using the **landscapemetrics** package
rsm = RootScapeMetrics( img, indexD = 5, metrics = c( "lsm_c_ca","lsm_l_ent","lsm_c_pd","lsm_c_enn_mn"))
```



