---
title: "Minirhizotron_Scans"
author: "Johannes Cunow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minirhizotron_Scans}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Minirhizotron Scans Analysis

## Introduction

This vignette demonstrates how to analyze minirhizotron scans using the **RootScanR Package**. This package is designed for analyzing root systems from images, including stitching, segmentation recommendations, leveraging in-situ calibration to create depthmaps, skeletonization, and extracting root traits and others.

## Key Functionalities

The package includes several primary functionalities:

1. **Image Stitching**
2. **Recommendations for AI Image Segmentation**
3. **Calibration Options**
4. **Depthmap Creation and Binning**
5. **Skeletonization and Root Trait Extraction**
6. **Extra Features** (e.g., peat texture, rhizosphere analysis)

## Installation

To install the package from the source or GitHub:

```{r}
# Install the package from GitHub
# install.packages("remotes")
# remotes::install_github("jcunow/RootScanR")

# install.packages("devtools")
# devtools::install_github("jcunow/RootScanR")

# Load the package
library(RootScanR)
```


## Example Workflow

### 1. Image Stitching 

If you use longer Minirhizotron tubes, you will need to stitch multiple images together into a single composite image for further analysis. This will remove redundant overlap between images and fascillitate true depth assignemnt:

```{r}
library(RootScanR)
library(tidyverse)
# Load example image files
image_files <- c("image1.png", "image2.png", "image3.png")

# Stitch images together
stitched_image <- stitch_sequential_images(image_files)  

# Visualize the stitched image
plot(stitched_image, main = "Stitched Image")
```

### 2 Recommendations for AI Image Segmentation (Minirhizotron Scans)

To segment roots from the background, we recommend using tools like **RootDetector** or **RootPainter**, which utilize AI for Minirhizotron images. Image segmentation is critical for any further processing. There are several options available: 
Distinguishing between root and background is essential. Two potential choices:

1. “RootDetector” applied to stitched Images;
   This software will return a segmented and skeletonized Image with information stored in different channels. RootDetector allows to distinguish tape, roots, and background using the root identification method.    
   Peters et al. (2023) As good as but much more efficient and reproducible 
   than human experts in detecting plant roots in minirhizotron images: 
   The Convolution Neural Network RootDetector  *Scientific Reports* **13** 1,

   Using root tracking will return root production, root decay, and no-change roots. Using this unlocks a additional Turnover estimation - Turnover.DPC()
   Gillert et al. 2023, Tracking Growth and Decay of Plant Roots in Minirhizotron Images. *IEEE/CVF Winter Conference on Applications of Computer Vision (WACV)*, https://doi.org/10.1109/WACV56688.2023.00369 
 
   Software at https://github.com/ExPlEcoGreifswald/RootDetector

2. "RootPainter" applied to stitched Images;
   This software returns a segmented Image.   
   Smith et al. 2022, RootPainter: deep learning segmentation of biological images with corrective annotation, *New Phytologist* **236** 2, https://doi.org/10.1111/nph.18387
   
   Software at https://github.com/Abe404/root_painter 

Tipp: split your stitched images if the root detection algorithm has trouble with the amount of roots of roots. You can rejoin the segmented images afterwards. But first, look if the settings can be changed to crank up the maximum root limit.  

### 3. Load Images

```{r}

img = terra::rast(seg_Oulanka2023_Session01_T067)
rgb.img = terra::rast(rgb_Oulanka2023_Session03_T067)
```



### 3. Calibration

If you use MinirhizotronTubes, accurate depth attributation requires a calibration of the tubes in-situ, see [link]. It's important to know how the scans relate to soil depth and the location of where tube upside face in the image. If we don't have *in-situ* calibration available, we can try to make a gross estimate. We are going to give our best guess on soil surface position and the rotation center position by using tape cover position as proxy.

```{r}
# In-situ calibration is highly highly recommended. 
# If you don't have any calibration data available - too bad!
# Try this instead:
# Based on Tape Cover, soil start is estimated
s0 = RootScanR::SoilSurfE(rgb.img)
print(paste0("Soil Surface Beginning in columns: ",s0$soil0))

# While we're at it, lets estimate the rotion center (Here, we assume that the tape extrusion on the upside of the Tube is perfectly centered - go and calibrate !!!)
r0 = RootScanR::RotationE(rgb.img)
print(paste0("Rotation Center in rows: ",r0))

```

### 4. Depthmap Creation and Binning
We are ready now to create a depth map. The goal here is to supply every pixel of the root scans with a depth information. We can choose to include the tube thickness and thereby create a phase shifted cosine depth map  -  or ignore it  (as traditionally done). Ignoring it will create a map in which the tube upside is located the same depth as tube downside. 


#### Step 1: Create Depthmap

```{r}

# We want to exclude pixels with tape as part of the soil. Create a mask:
# The format of 'RootDetector' uses the red channel to include the tape, while blue and green only show root pixel. Subtraction leaves as with the tape only.
mask = (seg_Oulanka2023_Session03_T067[[1]] - seg_Oulanka2023_Session03_T067[[2]] ) / 255
mask = terra::t(mask)

# Create the Map (soil0 = image above the soil surface (px); r0 = rotation center (0-1))
center.offset = r0  / dim(seg_Oulanka2023_Session03_T067)[1]
DepthMap = create.depthmap(im = seg_Oulanka2023_Session03_T067, 
                           sinoid = TRUE, 
                           mask = mask,start.soil = s0,
                           center.offset = center.offset )
```

#### Step 2: Bin the Depthmap
# Now lets bin the depth to be able to calculate some stats.We use 5cm intervals here but any bandwidth can be used.
```{r}
# nn controls depth band width
bm = RootScanR::binning(DepthMap, nn = 5, "rounding")
terra::ext(bm) <- c(0,dim(bm)[2],0,dim(bm)[1])
# Binned Depthmap Plot
tbm = terra::t(bm)
terra::plot(tbm)

```


### 5. Skeletonization
# Many operations will require having a centerline skeleton of the roots.
```{r}
skeleton <- skeletonize_image(img, verbose = TRUE, methods = "MAT", select.layer = 2)
plot(skeleton, main = "Skeletonized Image")
```


### 6. Root Trait Extraction
# The RootScanR package offers a couple of root trait estimations. 

```{r}
# Extract Brancnhing Points
SkeletonPoints <- detect_skeleton_points(skeleton)
roottips = px.sum(SkeletonPoints$Endpoints)
branchingpoints = px.sum(SkeletonPoints$Branchingoints)

# Get Kimura Root Length
rootlength = RootLength(skeleton)

# Calculate Root Cover Density
void = (load_flexible_image(img, binarize = TRUE)-1)*-1
rootpx = px.sum(img, select.layer = 2)
voidpx = px.sum(void, select.layer = 2)
rootdensity = rootpx / (rootpx + voidpx)

# Get the 'deepdrive' trait
ddrive = deep.drive(RootMap = img, DepthMap = DepthMap)

# a couple of others using the **landscapemetrics** package
rsm = RootScanRMetrics( img, indexD = 5, metrics = c( "lsm_c_ca","lsm_l_ent","lsm_c_pd","lsm_c_enn_mn"))
```



### 7. Soil Characterization
```{r}
# The overal color
Tube.color = Tube.coloration(rgb.img)

# Root density and peat color can be correlated. Lets try with only soil. First,lets create a buffer around root pixels.
buff.im = Halo(img, width = 2, halo.only = FALSE)

# substract the buffer from the rgb image
peat.img = rgb.img - buff.im

# Now we can look at the color without roots
Peat.color = Tube.coloration(peat.im)

# We can also try to extract information of the optical texture of the peat
gray.img = raster::raster(rgb2gray(peat.img))
void.tx = glcm::glcm(gray.img,
                     n_grey = 7,
                     window = c(3,3), 
                     statistics =
                       c("second_moment","homogeneity"),
                     shift=
                       list(c(0,1), c(1,1), c(1,0), c(1,-1)))



```



### 7. Depth Slice Analysis

```{r}


```
