---
title: "MinirhizotronR: Depth Resolved Analysis of Segmented Minirhizotron Scans"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(include = FALSE)
library(MinirhizotronR)
```

## Description: 
This package is designed in support a analysis of root scans from Minirhizotron tubes, starting after image segmentation.
To use this package to its full potential, you will need :
- mandatory: raw root scans and the segmented equivalent. 
- optional: skeletonized equivalents, in-situ calibration of soil surface to image start, as well as rotation calibration is beneficial
- suggestion: 
- use 'RootDetector' or an equivalent Image segmentation convolutional neural net to aid in automated analysis of root scans beforehand.
- use AffineStitcher.py or equivalent to stitch all individual scans belonging to a particular tube in time to generate a unified image
    
## Example Show Case     
```{r}
# Root per Available Space ~ Depth + Plant Community

```

## Install Package
```{r}
devtools::install_git("jcunow/MinirhizotronR")
```

